Efficient inference, potential, and limitations of site-specific substitution models

### 2020-07-26 manuscript/response_to_referees.tex

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1}
\newcommand{\refb}[1]{\textbf{R2:} #1}
\newcommand{\editor}[1]{\textbf{Editor:} #1}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\color{black}\textbf{Response:} #1}}

\begin{document}
\section*{Response to reviewers}

\subsection*{Associate Editor: Martin, Darren}

Thank you very much for handling the paper and providing these two constructive reviews that have helped us to improve the paper. Please accept our apologies for the delayed resubmission. My two co-authors have moved one (more than a year ago) and I have been caught up in SARS-CoV-2 related work for much of the year.

We realized that our original manuscript was misleading to people familiar complex mixture models that partition the sequence data into categories with similar properties. We now make it very explicit that this is not what we study -- instead, we study the extent to which we can estimate specific properties of every column of an alignment. Many of the comments by Reviewer 2 suggest that they were expecting an analysis of partition models while in fact, we studied the identifiability of parameters describing selection at every single site in the sequence. Such inferences can complement deep-mutational scanning techniques and become increasingly feasible as more and more diverse sequences of RNA viruses are available. We, therefore, think that the paper is of interest to the readership of Virus Evolution.

@@

\editor{This seems like a good paper but in its current state virus evolution may not be a very good fit for it.  I think though that if you follow reviewer 2's recommendations (especially with respect to providing practical advice to virologists) it should be relatively straightforward to make the paper a bit more accessible to virus evolution's readership.}

\response{
We have rewritten large parts of the manuscript to make our work more accessible and relevant to the readership of Virus Evolution.
Since our primary objective is to learn about the biology of virus from sequence diversity, rather than develop methods for phylogenetic inference, we believe Virus Evolution and not Systematic Biology is an appropriate venue for this work.
The referee reports were very useful in rewriting and refocussing our work.
}

@@

\refa{- Fig 5 and Fig S4 are both over-plotted and making the points transparent or adding histograms would help with interpretability. In Fig 1, the differences in linetype should be included in the legend. In Fig 2, it would be easier to keep the models straight if the lines were annotated on the graph directly instead of relying exclusively on color.}

\response{
We are grateful for these suggestions to improve the figures:
\begin{itemize}
    \item we changed Fig 5 and Fig S4 to two dimensional histograms
    \item we added second legend to figure 1 that maps line style to samples size. We also removed the curves corresponding to naive estimates by alignment frequencies. These are also include in Fig 2 and hence not needed here.
    \item labeling individual lines in Fig 2 proved a bit tricky, but we added markers to the lines such they can be distinguished without reference to color.
\end{itemize}
}

\subsection*{Reviewer: 2}

We thank the reviewer for their detailed review of our manuscript and the many constructive comments.
This review as revealed that we should have been much more explicit about the relation of our manuscript to partition models commonly used to model site-heterogeneity of rates or preferences.
Instead of partitioning the sequence data into a small number of categories, we investigate to what extent we can learn parameters that describe specific selective pressures at every column in the alignment.
The primary objective is to learn about the function of the sequence, not phylogenetic inference.

### 2020-07-27 manuscript/response_to_referees.tex

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1}
\newcommand{\refb}[1]{\textbf{R2:} #1}
\newcommand{\editor}[1]{\textbf{Editor:} #1}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}}

@@

\begin{document}
\section*{Response to reviewers}

\subsection*{Associate Editor: Martin, Darren}

\response{
Thank you very much for handling our manuscript and providing these two constructive reviews that have helped us to improve the paper. Please accept our apologies for the delayed resubmission. My two co-authors have moved one (more than a year ago) and I have been caught up in SARS-CoV-2 related work for much of the year.

We realized that our original manuscript was misleading to people familiar with complex mixture models that partition the sequence data into categories with similar properties. We now make it very explicit that this is not what we study -- instead, we study the extent to which we can estimate specific properties of every column of an alignment. Many of the comments by Reviewer 2 suggest that they were expecting an analysis of partition models while in fact, we studied the identifiability of parameters describing selection at every single site in the sequence. Such inferences can complement deep-mutational scanning techniques and become increasingly feasible as more and more diverse sequences of RNA viruses are available. We, therefore, think that the paper is of interest to the readership of Virus Evolution.
}

@@

\refa{Puller, Sagulenko, and Neher present a systematic look at the strengths and weaknesses of site-specific substitution models. They propose a new method for inferring site-specific parameters, investigate the limitations of the model, and applied it to within-in host and global HIV sequences. This work is clear, methodical, and will be of great interest to those who work on site-specific models. I recommend this work for publication. I have also listed some comments below but do not see any of them critical to acceptance.}

\response{Thank you very much for this assessment.}

\refa{- The authors mention that other methods to infer site-specific parameters do exist. How does their new algorithm compare in terms of speed or accuracy?}

\response{The most obvious software to compare our results to would be \texttt{swmutsel}.
However, \texttt{swmutsel} specifically models sequences are amino acides and their codons, something that isn't currenly implemented in treetime. In terms of run-time, swmutsel takes hours where the site specific inference by treetime takes minutes.
But we are afraid the results are not immediately comparable.
We have hence refrained from attempting a comparision.
}

@@

\refa{- For the simulations, the author draw site-specific preferences from a Dirichlet distribution with $\alpha$ equal to 0.2 or 0.5. It would be nice to present a more interpretable value of the preferences as well, such as entropy or effective number.}

\response{
Thank you for this suggestion.
For amino acid alphabets, $\alpha=0.2$ and $\alpha=0.5$ correspond to an average entropy of $1.8$ or $2.3$ (in natural log units), respectively.
Calculating the effective number of states as $\left(\sum_i p_i^2\right)^{-1}$ yields 4.7 or 7.8 effective states.
We added th the latter to the manuscript.}

\refa{- Fig 5 and Fig S4 are both over-plotted and making the points transparent or adding histograms would help with interpretability. In Fig 1, the differences in linetype should be included in the legend. In Fig 2, it would be easier to keep the models straight if the lines were annotated on the graph directly instead of relying exclusively on color.}

\response{
We are grateful for these suggestions to improve the figures:
\begin{itemize}
    \item we reduced marker size and opacity in Fig 5 and Fig S4 to reduce overplotting. We also experimented with two dimensional histograms, but bins at the border of the domain (almost invariable sites) dominate to a degree that it is hard to see differences in the remainder of the histogram.
    \item we added second legend to figure 1 that maps line style to samples size. We also removed the curves corresponding to naive estimates by alignment frequencies. These are also include in Fig 2 and hence not needed here.
    \item labeling individual lines in Fig 2 proved a bit tricky, but we added markers to the lines such they can be distinguished without reference to color.
\end{itemize}
}

\subsection*{Reviewer: 2}
\response{
We thank the reviewer for their detailed review of our manuscript and the many constructive comments.
This review as revealed that we should have been much more explicit about the relation of our manuscript to partition models commonly used to model site-heterogeneity of rates or preferences.
Instead of partitioning the sequence data into a small number of categories, we investigate to what extent we can learn parameters that describe specific selective pressures at every column in the alignment.
The primary objective is to learn about the function of the sequence, not phylogenetic inference.
}

@@

\refb{Abstract, lines 24-25: I assume it will be pointed out that the BEAST/BEAST2 packages do allow for site-specific models to be used (with for example a simple button that allows to select the SRD06 model, of which the paper should be cited)?}

\response{We now discuss site-heterogenous partition models carefully in the main text.}

\refb{Lines 120-122: This brings to mind a recent paper by Gascuel and Steel ‘A Darwinian Uncertainty Principle’ in which the issue of jointly estimating the ancestral states and the transitions of those states is investigated. Given the parameters the authors are interested in here and their assessment through simulations, it seems to me that the authors need to clarify the novelty of their investigations and how theirs sets them apart from those investigated by Gascuel and Steel (actually Figure 1 reminds me of Figure 2 in the paper by Gascuel and Steel).}

\response{The paper by Gascuel and Steel (which we were unaware of) discusses indeed similar topics as we do. Gascuel and Steel investigate whether the state of the root of a character can be inferred jointly with the equilibrium probabilities or the transition matrix. Consistent with our results, they find that this is possible for Yule trees.
Here, we are mostly concerned with estimating the model, not the root state, and treat ancestral states as nuissance parameters that at least in principle can be integrated out.
In constrast to Gascuel and Steel, we consider sequence evolution rather than the evolution of a single character which allows our model to share the symmetric transmission matrix across sites.
It also allows us to investigate the extent to which branch lengths can be estimated along with the model.

We now discuss our results in the light of the results by Gascuel and Steel and thank the reviewer for pointing out this work to us.
G\&S's work is mentioned in the section discussing the accuracy of inferences and the discussion.
}

@@

\response{In essence, Fig 3 shows that branch length underestimation caused by extensive variation of rates and preferences is very hard to avoid.
As outlined above, we don't really have a good fix for that, but we investigate the causes in Fig 4.
That said, the main result of the paper is that site-specific frequencies can be accurately estimated.
The inability is estimating the length of long branches in RNA viruses is a serious concern for dating zoonosis and exploring the history of viruses causing measles, dengue, lassa etc.
This problem is clearly apparent when trying to date the common ancestor of SIVs.
Molecular clock approaches suggest an ancestor a few hundred years in the past, while viruses in island populations show that the common ancestor is much more ancient.
}

\refb{The manuscript deals with estimates (i.e. trees) that were obtained using IQ-TREE and FastTree (which produce unrooted trees and don’t employ the concept of a molecular clock), which were subsequently fed into TreeTime which employs the provided trees as is. This makes me wonder if the conclusions from the manuscript hold up in frameworks that i) employ a molecular clock directly in their estimations, and ii) do not employ a fixed topology but rather explore a distribution of trees as typically done in Bayesian phylogenetic inference. I’m not suggesting the authors do the same experiments in a Bayesian framework but their take on this could be interesting for the discussion section.}

\response{
We are afraid there is a misunderstanding here.
Our manuscript infers GTR models and the root does not matter.
We use root-to-tip distances in mid-point-rooted trees as a measure of divergence, but the root is no special place otherwise.
The only place where time comes in are the HIV examples, but we only use time as a proxy for divergence via the known evolutionary rate and don't do any molecular clock analysis.
All we want to do is to point out that branch length under-estimation potentially has consequences for dating rapidly evolving RNA viruses.}
